---
title: "SR_Lec06_notes"
format: html
editor: visual
---

## Lec 6: Good and Bad Controls

```{r}
#| output: false
#| error: false
#| warning: false
#| message: false
library(rethinking)
library(dagitty)
```

**Backdoor Criterion**: Rule to find a set of variables to stratify by to yield $P(Y|do(X))$

1.  Identify all **paths** connecting the treatment X to the outcome Y
    1.  The causes follow arrows but associations go both ways
2.  Paths with arrows **entering** X are backdoor paths (non-causal paths; confounding path)
    1.  if you intervene on X, the change will not flow into backdoor path
    2.  only arrows leaving X will be affected when changing X
3.  Find **adjustment set** that closes/blocks all backdoor paths

#### Example Flow

```{r}
#| echo: false
dag_e1 <- dagitty( "dag {
  U [unobserved]
  X -> Y
  X <- Z 
  Z <- U 
  U -> Y 
}")
coordinates( dag_e1 ) <- list( x=c(X=0, Z=0.25, U=0.5,Y=1) ,
                              y=c(X=1, Z= 0.5, U=0, Y=1) )
drawdag( dag_e1 )
```

1\) Identify all paths

- $X→ Y$

- $X ← Z ← U → Y$

2\) Identify backdoor paths

- $X ← Z ← U → Y$ is a backdoor path

3\) Minimal adjustment set

> Block the pipe by stratifying by Z: X \_ \|\| \_ U \| Z

$$P(Y|do(X)) = \sum_z P(Y|X,Z) P(Z=z)$$

$$Y_i \sim {\sf Normal}(\mu_i,\sigma)$$

$$\mu_i = \alpha + \beta_XX_i + \beta_ZZ_i$$

```{r}
# simulate confounded Y 
N <- 200 
b_XY <- 0 
b_UY <- -1
b_UZ <- -1
b_ZX <- 1 

set.seed(10)
U <- rbern(N)
Z <- rnorm(N, b_UZ*U)
X <- rnorm(N, b_ZX*Z)
Y <- rnorm(N, b_XY*X + b_UY*U)
d <- list(Y=Y, X=X, Z=Z)

# ignore U and Z (to compare first path vs second) 
m_YX <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + b_XY*X,
    a ~ dnorm(0,1),
    b_XY ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=d
)

# Stratify by Z 
m_YXZ <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + b_XY*X +b_Z*Z,
    a ~ dnorm(0,1),
    b_XY ~ dnorm(0,1),
    b_Z ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=d
)

post <- extract.samples(m_YX)
post2 <- extract.samples(m_YXZ)
dens(post$b_XY, lwd=3, col=1, xlab="posterior b_XY", xlim=c(-0.3,0.3))
dens(post2$b_XY, lwd=3, col=2, add=TRUE)

```

- Remember, the true effect of X on Y is 0.

```{r}
precis(m_YXZ)
```

- in the model `m_YXZ` the coefficient for `b_Z` is [meaningless]{.underline} (it does NOT mean that 0.24 is coefficient for Z on Y)

  - It's a control variable - fallacy to intepret the coefficients of control variables!!

> For any variable you add to the model as part of the adjustment set (in order to make it possible to get an estimate of some other variable's coefficient) - usually its coefficients are **not interpretabl**e (Table 2 Fallacy)

#### More complicated example

```{r}
#| echo: false
dag_e2 <- dagitty( "dag {
  X -> Y
  X <- A -> Z <- B -> Y
  X <- Z -> Y
  X <- C -> Y
}")
coordinates( dag_e2 ) <- list( x=c(A=0, Z=0.5, B=1, X=0, Y=1, C=0.5) ,
                                 y=c(A=0, Z=0, B=0, X=1, Y=1, C=1.5) )
drawdag( dag_e2 )
```

**1) List all the paths connecting X and Y. Which need to be closed to estimate the effect of X on Y?**

$X → Y$: direct causal path, necessarily open

Try to find mathematical way to delete the arrows that are entering X

1.  $X ← C → Y$: non-causal, open
    1.  Close by stratifying by $C$ (add $C$ to adjustment set)
2.  $X ← Z → Y$: non-causal, open
    1.  Close by stratifying by $Z$; add $Z$ to adjustment set
3.  $X ←  A → Z ← B → Y$: non-causal, [closed]{.underline} (opened by stratifying on Z)
    1.  Use A and/or B to close
4.  $X ← A → Z → Y$: non-causal, open
    1.  Already closed because we are stratifying by $Z$
5.  $X ← Z ← B → Y$: non-causal, open
    1.  Effectively closed because we are stratifying by $Z$ (existing adjustment set)

> Minimum adjustment set: C, Z, and *either A or B*. (B is better choice)

- B is more statistically efficient than A

#### Example

```{r}
#| echo: false
dag_e3 <- dagitty("dag {
  U [unobserved]
  G -> P -> C
  G -> C
  P <- U -> C}"
)

coordinates(dag_e3) <- list(x=c(G=0, P=1,C=1, U=1.5), 
                            y=c(G=0,P=0, C=1,U=0.5))
drawdag(dag_e3)
```

G: grandparent education\
P: parent education\
C: child education\
U: unobserved confound

1\)

1.  $G→ C$
2.  $G → P → C$: Pipe (P is a mediator)
3.  $G → P ← U → C$: P is a collider

To find the direct causal effect of $G$ on $C$, we need to stratify by $P$ (close the pipe). However, if we do that, then we open up Path 3 $G → P ← U → C$, defined by the fork $C← U → P$ (e.g., neighborhood variable). Can't get the direct causal effect of G on C because we activate the confounds. It is possible to calculate the total effect of G on C, but that isn't our research question.

+----------------------------------------+-----------------------------------------------------------------------+
| Can estimate TOTAL effect of G on C    | Cannot estimate the DIRECT effect                                     |
+========================================+=======================================================================+
| $C_i \sim {\sf Normal}(\mu_i, \sigma)$ | $C_i \sim {\sf Normal}(\mu_i, \sigma)$                                |
|                                        |                                                                       |
| $\mu_i = \alpha + \beta_G G_i$         | $\mu_i = \alpha + \beta_G G_i + \beta_PP_i$                           |
+----------------------------------------+-----------------------------------------------------------------------+
|                                        | Confounded estimate, may end up concluding that G hurts C's education |
+----------------------------------------+-----------------------------------------------------------------------+

## Controls

**control variable:** variable introduced to an analysis so that a causal estimate is possible

Wrong heuristics for choosing control variables:

- anything goes

- any variables not highly **collinear**

- Any pre-treatment measurement (baseline)

```{r}
#| echo: false
dag_e4 <- dagitty("dag {
  u [unobserved]
  v [unobserved]
  X -> Y 
  X <- u -> Z <- v -> Y
  } ")

coordinates(dag_e4) <- list(x=c(X=0, u=0, Z=0.5, v=1, Y=1),
                            y=c(X=1,u=0, Z=0.5, v=0, Y=1))
drawdag(dag_e4)
```

1.  $X → Y$ frontdoor and closed
2.  $X ← u → Z ← v → Y$ backdoor and closed
    1.  $Z$ is collider
    2.  Remains closed as long as we don't condition on $Z$
    3.  Not safe

```{r}
#| echo: false
dag_e5 <- dagitty("dag {
                  u [unobserved]
                  X -> Z -> Y 
                  Z <- u -> Y
}")

coordinates(dag_e5) <- list(x=c(X=0, Z=1, u=1.5, Y=2),
                            y=c(X=0.2, Z=0.2, u=0, Y=0.2))
drawdag(dag_e5)
```

X: win lottery\
Y: Lifespan\
Z: Happiness\
u: contextual confounds

Conditioning on $u$ is harmless UNLESS you try to stratify by $Z$! You can estimate the total causal effect of winning the lottery, but not direct causal effect.

- Post-treatment variables are often risky controls

1.  $X → Z → Y$

2.  $X → Z ← u → Y$

    ::: callout-note
    No backdoor path, no need to control for Z!!!
    :::

**Simulate the DAG**

Assume all arrows have effect of 1:

```{r}
f <- function(n=100, bXZ=1, bZY=1){
  X <- rnorm(n)
  u <- rnorm(n)
  Z <- rnorm(n, bXZ*X + u)
  Y <- rnorm(n, bZY*Z + u)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX,bXZ))
}

sim <- mcreplicate(1e4, f(), mc.cores = 8)

dens(sim[1,], lwd=3, xlab="posterior mean",xlim=c(-1,2), ylim=c(0,2.75))
dens(sim[2,], lwd=3, col=2, add=TRUE)
```

- Adding Z in `Y ~ X + Z` makes it look like the effect is negative because it activates the confound

- `Y ~ X` is correctly centered on `1.0`

Now, assume all are 1 except `bZY = 0` :

```{r}
sim <- mcreplicate(1e4, f(bZY=0), mc.cores = 8)

dens(sim[1,], lwd=3, xlab="posterior mean",xlim=c(-1,0.5), ylim=c(0,3))
dens(sim[2,], lwd=3, col=2, add=TRUE)
```

- Again, `Y ~ X` is correctly centered on 0 ($X$ has no causal influence on $Y$ )

- While `Y ~ X + Z` mistakenly shows a negative relationship

::: callout-note
Statistical significance is not a guide to model structure.
:::

#### Colliders

::: callout-note
Do not touch the collider! Colliders are BAD CONTROLS
:::

Colliders are not always obvious:

```{r}
#| echo: false
dag_e6 <- dagitty("dag {
                  u [unobserved]
                  X -> Z <- u -> Y
                  X -> Y
}")

coordinates(dag_e6) <- list(x=c(X=0, Z=0.5, u=1, Y=1),
                            y=c(X=1, Z=0, u=0, Y=1))
drawdag(dag_e6)
```

- Z becomes a collider if you consider the possibility of unmeasured causes between Z and Y

X: education\
Y: income\
Z: values\
u: family

EX: How does education affect income? Imagine you have data on people's values based on survey results, so you decide to condition on values (Z). But, you could be conditioning on a collider and accidentally get estimates that are higher SPURIOUS. Because values are often associated with family/upbringing (u), which can be associated with income (Y).

### Non-backdoor examples

#### Case-control bias (selection on the outcome)

$$X (education) → Y (occupation) → Z (Income) $$

- If you stratify based on income (Z), you're weakly stratifying on the occupation Y, the outcome itself. It narrows the range of occupations that experiences a narrower range of education levels (false inferences)

Let's see a simulation of this: (assume all arrows have coefficient of 1)

```{r}
f <- function(n=100, bXY=1, bYZ=1){
  X <- rnorm(n)
  Y <- rnorm(n, bXY*X)
  Z <- rnorm(n, bYZ*Y)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX,bXZ))
}

sim <- mcreplicate(1e4, f(), mc.cores = 8)

dens(sim[1,], lwd=3, xlab="posterior mean",xlim=c(0,1.5), ylim=c(0,5))
dens(sim[2,], lwd=3, col=2, add=TRUE)
```

- Conditioning on $Z$ gives a weaker association (by half)

#### Precision parasite

- Stratifying on the parent of X

  $$Z → X → Y $$

Simulate the DAG

```{r}
f <- function(n=100, bZX=1, bXY=1){
  Z <- rnorm(n)
  X <- rnorm(n, bZX*Z)
  Y <- rnorm(n, bXY*X)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX,bXZ))
}

sim <- mcreplicate(1e4, f(), mc.cores = 8)

dens(sim[1,], lwd=3, xlab="posterior mean")
dens(sim[2,], lwd=3, col=2, add=TRUE)
```

- Centers on the correct value, but it inflates the uncertainty (wider distribution)

#### Bias amplification

- X and Y confounded by u

- Something truly awful happens when we add Z

  - double activates the confound

```{r}
#| echo: false
dag_e7 <- dagitty("dag {
                  u [unobserved]
                  Z -> X <- u -> Y
                  X -> Y
}")

coordinates(dag_e7) <- list(x=c(X=0, Z=0, u=0.5, Y=1),
                            y=c(X=1, Z=0, u=0, Y=1))
drawdag(dag_e7)
```

X: occupation\
Y: income\
Z: education\
u: regional/cultural factors

EX: Examining the association between occupation and income. Lots of regional and cultural factors that influences both. Stratifying by education seems reasonable... but this is a bias amplification problem.

Simulate: assume bXY is 0

```{r}
f <- function(n=100, bZX=1, bXY=1){
  Z <- rnorm(n)
  u <- rnorm(n)
  X <- rnorm(n, bZX*Z + u)
  Y <- rnorm(n, bXY*X + u)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX,bXZ))
}

sim <- mcreplicate(1e4, f(bXY=0), mc.cores = 8)

dens(sim[1,], lwd=3, xlab="posterior mean")
dens(sim[2,], lwd=3, col=2, add=TRUE)
```

- Real value is 0 (no association between X and Y)... but the model finds association

- The default model is already biased (we knew that - just the effect of an unobserved confound)

- Conditioning on Z increases the bias

Why?

- Covariation between X and Y can only exist if there's variation in their causes

  - The cause of Y is X, and if X doesn't vary then Y can't vary

  - Then Z is a cause of X... so if Z doesn't vary then X can't vary

- Within each level of Z, there is less variation in X

  - When we stratify by Z, we remove the variation in Z that causes X

  - Therefore the covariation between X and Y must be smaller within each level of Z

- Consequently, confound $u$ becomes relatively more important within each Z (when we condition on Z)

  - When you average across the levels of Z, you find more association between X and Y

```{r}
n <- 1000 
Z <- rbern(n) # make Z binary to make viz of stratification easier (either 0 or 1)
u <- rnorm(n)
X <- rnorm(n, 7*Z + u)
Y <- rnorm(n, 0*X + u)

# looking for a slope of 0 = no association 

```

::: callout-note
Control variable: variable introduced to an analysis so that a causal estimate is possible

Heuristics fail – adding control variables can be worse than omitting

Make assumptions explicit (must do scientific modeling \[DAGs\] to do scientific data analysis)
:::
