---
title: "SR_Lec06_notes"
format: html
editor: visual
---

## Lec 6: Good and Bad Controls 

```{r}
library(rethinking)
library(dagitty)
```

**Backdoor Criterion**: Rule to find a set of variables to stratify by to yield $P(Y|do(X))$

1.  Identify all **paths** connecting the treatment X to the outcome Y
    1.  The causes follow arrows but associations go both ways
2.  Paths with arrows **entering** X are backdoor paths (non-causal paths; confounding path)
    1.  if you intervene on X, the change will not flow into backdoor path
    2.  only arrows leaving X will be affected when changing X
3.  Find **adjustment set** that closes/blocks all backdoor paths

#### Example Flow

```{r}
#| echo: false
dag_e1 <- dagitty( "dag {
  U [unobserved]
  X -> Y
  X <- Z 
  Z <- U 
  U -> Y 
}")
coordinates( dag_e1 ) <- list( x=c(X=0, Z=0.25, U=0.5,Y=1) ,
                              y=c(X=1, Z= 0.5, U=0, Y=1) )
drawdag( dag_e1 )
```

1\) Identify all paths

- $X→ Y$

- $X ← Z ← U → Y$

2\) Identify backdoor paths

- $X ← Z ← U → Y$ is a backdoor path

3\) Minimal adjustment set

> Block the pipe by stratifying by Z: X \_ \|\| \_ U \| Z

$$P(Y|do(X)) = \sum_z P(Y|X,Z) P(Z=z)$$

$$Y_i \sim {\sf Normal}(\mu_i,\sigma)$$

$$\mu_i = \alpha + \beta_XX_i + \beta_ZZ_i$$

```{r}
# simulate confounded Y 
N <- 200 
b_XY <- 0 
b_UY <- -1
b_UZ <- -1
b_ZX <- 1 

set.seed(10)
U <- rbern(N)
Z <- rnorm(N, b_UZ*U)
X <- rnorm(N, b_ZX*Z)
Y <- rnorm(N, b_XY*X + b_UY*U)
d <- list(Y=Y, X=X, Z=Z)

# ignore U and Z (to compare first path vs second) 
m_YX <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + b_XY*X,
    a ~ dnorm(0,1),
    b_XY ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=d
)

# Stratify by Z 
m_YXZ <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + b_XY*X +b_Z*Z,
    a ~ dnorm(0,1),
    b_XY ~ dnorm(0,1),
    b_Z ~ dnorm(0,1),
    sigma ~ dexp(1)
  ), data=d
)

post <- extract.samples(m_YX)
post2 <- extract.samples(m_YXZ)
dens(post$b_XY, lwd=3, col=1, xlab="posterior b_XY", xlim=c(-0.3,0.3))
dens(post2$b_XY, lwd=3, col=2, add=TRUE)

```

> For any variable you add to the model as part of the adjustment set (in order to make it possible to get an estimate of some other variable's coefficient) - usually its coefficients are not interpretable (Table 2 Fallacy)

1.  $X ←  A → Z ← B → Y$: non-causal, closed (opened by Z)
    1.  A or B to close
2.  $X ← A → Z → Y$: non-causal, open
3.  $X ← Z ← B → Y$: non-causal, open
4.  $X ← Z → Y$: non-causal, open
    1.  add $Z$ to adjustment set
5.  $X → Y$: causal path, open
6.  $X ← C → Y$: non-causal, closed
    1.  add $C$ to adjustment set

Try to find mathematical way to delete the arrows that are entering X

Minimum adjustment set: C, Z, and either A or B. (B is better choice)

#### 

No backdoor path, no need to control for Z

#### Colliders

Do not touch the collider! Colliders are BAD CONTROLS

### Non-backdoor examples

#### Case-control bias (selection on the outcome) 

- X (education) → Y (occupation) → Z (Income)

- If you stratify based on income, it narrows the range of occupations that experiences a narrower range of education levels (false inferences)

#### Precision parasite

- Stratifying on the parent of X

  - Z → X → Y

- inflates the uncertainty (wider distribution)

#### Bias amplification 

- X and Y confounded by u
