---
title: "Ch3: Sampling the Imaginary"
format: html
editor: visual
date: 04-11-2026
---

```{r setup, include=FALSE}
library(rethinking)
```

Common example when teaching Bayesian statistics intro:

```{r}
Pr_Positive_Vamp <- 0.95
Pr_Positive_mortal <- 0.01
Pr_Vamp <- 0.001
Pr_Positive <- Pr_Positive_Vamp * Pr_Vamp + Pr_Positive_mortal * (1- Pr_Vamp )

(Pr_Vamp_Positive <- (Pr_Positive_Vamp * Pr_Vamp)/Pr_Positive)
```

**Not intuitive**: Whenever the condition of interest is very rare, having a test that finds all the true cases is still no guarantee that a positive result carries much info at all. Reason being: most positive results - false positives.

Again, though, not inherently Bayesian - just uses the theorem. Since all the probabilities are about the frequencies of events, rather than theoretical parameters, all major stat philosophies agree to use Bayes' theorem in this case.

**Frequency format:** Posed in a way that's more intuitive..

1.  In a population of 100,000 people, 100 are vampires.
2.  Of the 100 vampires, 95 will test positive for vampirism.
3.  Of the 99,900 mortals, 999 will test positive for vampirism.

If we test all 100,000 people, what proportion of those who test positive for vampirism truly vampires? Now, just count up the positive testers: 95 + 999 = 1094. Of the 1094, 95 are real vamps, so

$$Pr(vampire|positive) = 95/1094 \approx 0.087$$

Exploiting frequency bias for predicting things with posterior distributions!

- Instead of integral calculus, we'll just sample parameter values from the probability distributions. This turns it into a problem of data summary (a frequency format).

  - Inside a computer, parameters are just as empirical as outcome of coin flip or die toss.

  - Empirically attack the posterior

::: callout-note
Scientific inference often framed:

1.  An hypothesis is either T or F
2.  We use statistical procedure and get imperfect cue of hypothesis' falsity
3.  We (should) use Bayes' Theorem to logically deduce the impact of the cue on our uncertainty

**Toy example to see how little statistical procedures can do for us:** Suppose the probability of a positive finding, when a hypothesis is true, is Pr(sig\|true) = 0.95. This is the *power* of the test. Suppose the probability of a positive finding, when a hypothesis is false, is Pr(sig\|false) = 0.05. That's the false-positive (Type II Error) rate, like the 5% of conventional significance testing. Finally, we have to state the *base rate* at which hypotheses are true. Suppose for example that 1 in every 100 hypotheses turn out to be true. Then Pr(true) = 0.01. No one knows this value, but the history of science suggests it's small (very few long-standing true hypotheses). Now use Bayes' to compute the posterior:

$$Pr(true|pos) = \frac{Pr(pos|true)Pr(true)}{Pr(pos)} = \frac{Pr(pos|true)Pr(true)}{Pr(pos|true)Pr(true) + Pr(pos|false)Pr(false)}$$

Answer is approx 0.16. A positive finding corresponds to true hypothesis only 16% of the time. Same low base-rate phenomenon that applies in medical testing. Can shrink the false-positive rate to 1% and get the posterior prob up to 0.5, only as good as coin flip.

The most important thing to do is improve Pr(true), the base rate... just come up with a better hypothesis = thinking, not testing!
:::

### Sampling from a grid-approximate posterior

```{r 3.2}
p_grid <- seq(0, 1, length.out = 1000)
prob_p <- rep(1, 1000)
prob_data <- dbinom(6, 9, prob=p_grid) #likelihood p^L * (1-p)^W
posterior <-  prob_data * prob_p #unstandardized posterior 
posterior <- posterior/sum(posterior) # standardized

plot(p_grid, posterior, type="l")
```

Draw 10,000 samples from this posterior. Imagine the posterior is a bucket filled with parameter values (the $p$s, like 0.25, 0.1, 0.93, 0.5, 1, etc.). Within the bucket, each value exists in proportion to its posterior probability, such that values near the peak are much more common than those in the tails. Scoop out 10,000 values. Provided the bucket is well mixed, the resulting samples will have the same proportions as the exact posterior density. Therefore, the individual values of $p$ will appear in our samples in proportion to the posterior plausibility of each value.

`sample` pulls values from vector randomly.

```{r 3.3}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
plot(samples)
```

In this plot, it's like you're flying over the posterior distribution, looking down at it. There are many more samples from the dense region near 0.6 and few samples below 0.25.

```{r 3.5}
dens(samples)
```

`dens` function from the `rethinking` library computes the density estimate from the samples and plots it against the proportion of water (p).

Estimated density is similar to ideal posterior in 3.2.

### Sampling to summarize

Now, use samples to describe and understand the posterior. How it's summarized depends on research goals. Common questions:

- How much posterior probability lies below some parameter value?

- How much posterior probability lies between two parameter values?

- Which parameter value marks the lower 5% of the posterior probability?

- Which range of parameter values contains 90% of the posterior prob?

- Which parameter value has the highest posterior probability?

Three types of questions: (1) intervals of defined boundaries, (2) Intervals of defined probability mass, and (3) Point estimates.

#### (1) Intervals of defined boundaries

What's the posterior prob that the proportion of water is less than 50%?

Ok, without using the samples, we can easily calculate this with the grid-approx posterior:

```{r 3.6}
sum(posterior[p_grid < 0.5])
```

This is easy with one parameter examples. But if there's more parameters, will get more complicated. That's when we'd use the sampling method to describe the data. So take the sum of sampled parameter values that are less that 0.5 divided by 10000 (the frequency).

```{r 3.7}
sum(samples < 0.5) / 1e4
```

Also get how much post prob lies between 0.5 and 0.75:

```{r 3.8}
sum(samples > 0.5 & samples < 0.75) / 1e4

```

#### (2) Intervals of defined mass

Getting the integral. Usually known as **confidence interval**. He calls it **Compatibility interval**. it is a range of parameter values compatible with the model and data.

Easier to do this with sampling method than the computed grid-approximate posterior distribution.

Find boundaries of lower 80% posterior probability:

```{r 3.9}
quantile(samples, 0.8)
```

Middle 80% interval is between the 10th percentile and 90th percentile:

```{r 3.10}
quantile(samples, c(0.1,0.9))
```

**Percentile intervals**: communicate the shape of distribution, given that the distribution is not too asymmetrical. But imperfect when trying to support inferences about which parameters are consistent with data.

Consider observing 3 Ws in 3 tosses and uniform(flat) prior:

```{r 3.11}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(3, 3, prob=p_grid)
posterior <- prior * likelihood 
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, size=1e4, replace=TRUE, prob=posterior)

PI(samples, prob=0.5)
```

`PI(samples, prob)` from the `rethinking` package computes the central percentile compatibility interval

Highly skewed, its max value is at the boundary of p=1 (right end of x-axis). The percentile interval misses the skewed shape of the posterior.

In contrast, use the **highest posterior density interval (HPDI)**: the narrowest interval containing the specified probability mass. Want the interval with the parameter vals most consistent with the data = densest of intervals.

```{r 3.13}
HPDI(samples, prob = 0.5)
```

This interval captures the parameters w/ highest posterior prob, and is noticeably narrower: 0.16 in width compared to the 0.23 for the PI.

**Benefits of HPDI is mostly seen when the posterior is highly skewed.** When the posterior is bell shaped, it shouldn't matter which interval type you use.

```{r}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, 9, prob=p_grid)
posterior <- prior * likelihood 
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, size=1e4, replace=TRUE, prob=posterior)

PI(samples, prob=0.5)
```

```{r}
HPDI(samples, prob = 0.5)
```

**Disadvantages of HPDI**: more computationally intensive, suffers greater simulation variance (sensitive to how many samples are drawn from the posterior)

::: {.callout-note style="color: slategray;"}
Overall, if the choice of interval type makes a big difference, then you shouldn’t be using intervals to summarize the posterior. **Remember, the entire posterior distribution is the Bayesian “estimate.”** It summarizes the relative plausibilities of each possible value of the parameter. **Intervals of the distribution are just helpful for summarizing it**. If choice of interval leads to different inferences, then you’d be better off just plotting the entire posterior distribution.
:::

#### (3) Point estimates

Bayesian estimate is the entire posterior distribution. Producing point estimates often discards information.

Can get the highest post probability - maximum a posteriori (MAP) estimate. Using 3W in 3 tosses example:

```{r}
p_grid[which.max(posterior)]
```

Or from the samples from the posterior:

```{r}
chainmode(samples, adj=0.01)
```

Why mode verses posterior mean or median?

```{r}
mean(samples)
median(samples)
```

**Loss function**: rule that displays the cost associated with using any particular point estimate

- Different loss functions imply different point estimates

*Heuristic vignette:* Tell me which value of p, the proportion of water on the Earth, you think is correct. I will pay you \$100, if you get it exactly right. But I will subtract money from your gain, proportional to the distance of your decision from the correct value. Precisely, your loss is proportional to the absolute value of d−p, where d is your decision and p is the correct answer.

- We could change the precise dollar values involved, without changing the important aspects of this problem. What matters is that the loss is proportional to the distance of your decision from the true value.

Calculating expected loss for any given decision - use posterior to average over our uncertainty in the true value. Don't know the true value, so use entire posterior distribution.

Suppose decision, p=0.5. Expected loss will be

```{r 3.17}
sum(posterior * abs(0.5 - p_grid))
```

Compute the weighted average loss, where each loss is weighted by its corresponding posterior probability.

Trick to compute for every possible decision: `sapply`

```{r 3.18}
loss <- sapply(p_grid, function(d) sum(posterior * abs(d- p_grid)))
# loss contains all possible decision loss values. we want to find the value that minimizes the loss
p_grid[which.min(loss)] # gets the index of the min value, finds correspoding val on p_grid
```

Which is the *posterior median*: the parameter value that splits the posterior density such that half the mass is above it and below it.

```{r}
median(samples)
```

Different loss functions nominate diff point estimates. Two common ones:

1.  Absolute loss - median as point estimate
2.  Quadratic loss (d-p)\^2 - posterior mean

When posterior distribution is symmetrical and normal-looking - mean and median converge

::: {.callout-note style="color: slategray;"}
Usually, research scientists don’t think about loss functions. And so any point estimate like the mean or MAP that they may report isn’t intended to support any particular decision, but rather to describe the shape of the posterior. You might argue that the decision to make is whether or not to accept an hypothesis. But the challenge then is to say what the relevant costs and benefits would be, in terms of the knowledge gained or lost. **Usually it’s better to communicate as much as you can about the posterior distribution, as well as the data and the model itself, so that others can build upon your work**. Premature decisions to accept or reject hypotheses can cost lives.
:::

## Sampling to simulate prediction

Likelihood functions work in both directions: given a realized obs, the likelihood function says how plausible the obs is. Given only parameters, the likelihood defines a distribution of possible obs that we can sample from to simulate obs. = generative models simulate predictions

**dummy data**: simulated data in generative models

Globe-tossing example. Two tosses: three possible observations (0 W, 1W, 2W). Compute probability for each, for any value of prior p.

```{r 3.20}
# p = 0.7 
dbinom(0:2, size=2, prob=0.7)
```

0W: 9%; 1W: 42%, 2W: 49%

Simulate observations using these probabilities.

```{r 3.21}
# r for random binomial, (# of simulations, size of N tosses, prob)
rbinom(1, size=2, prob=0.7)
rbinom(10, size=2, prob=0.7)
```

```{r 2.23}
dummy_w <- rbinom(100000, size=2, prob=0.7)
table(dummy_w)/1e5
```

```{r 3.24}
dummy_w <- rbinom(1e5, size=6, prob=0.7)
simplehist(dummy_w, xlab = "dummy water count")
```

**Model checking**: (1) ensuring the model fitting worked correctly, (2) evaluating the adequacy of a model for some purpose

1.  Software working correctly? Retrodictions ask how well the model reproduces the data used to educate it
2.  Model adequacy? All models are false, so we're not testing whether the "truthiness" of the model;s assumptions. We're trying to assess exactly how the model fails to describe the data, as a path toward model comprehension, revision, and improvement
    1.  Typically hope to either predict future observations or understand enough that we might usefully tinker with the world

Implied predictions in models are uncertain in two ways:

1.  Observation uncertainty: for any unique value of the parameter *p*, there is a unique implied pattern of observation the model expects (garden of forking data). Uncertainty in predicted observations, because even if you know p with certainty, you won't know the next globe toss with certainty (unless p=0 or 1).
2.  Prior uncertainty, embodied by the posterior distribution. Everything depends on p - so uncertainty extends outward, affecting the sampling variation.
    1.  Need to *propagate* the parameter uncertainty when evaluating implied predictions. So over the posterior density for p, compute predictions for each value of p. - average all these prediction distributions together to get the **posterior predictive distribution**.

```{r 3.25}
w <- rbinom(1e4, size=9, prob=0.6)
simplehist(w)
```

This generates 1e4 simulated predictions of 9 globe tosses assuming p = 0.6. The predictions are stored as counts of water. Theoretical min = 0; max = 9.

To propagate the parameter uncertainty into these predictions: replace the value 0.6 with samples from the posterior

```{r}
w <- rbinom(1e4, size=9, prob=samples)
simplehist(w)
```

`samples` is the same list of random samples from the posterior distributions used before. For each sampled value, a random binomial observation is generated.

- Combine simulated observation distributions for all parameter values, each weighted by its posterior prob → posterior predictive distribution

  - Propagates uncertainty about parameter to uncertainty about prediction

- Can manipulate these simulated observations just like you can manipulate samples from the posterior (compute intervals, point estimates)

The simulated model predictions are quite consistent with the observed data in this case - the actual count of 6 lies right in the middle of the simulated distribution. Quite a bit of spread to predictions, but this spread arises from the binomial process itself, not uncertainty about p. Still premature to conclude the model is perfect. Questionable assumption in the independence of globe tosses.
