---
title: "SR_Ch04_notes"
format: html
editor: visual
---

```{r setup, include=FALSE}
library(rethinking)
data(Howell1)
```

# Ch 4: Geocentric Models Notes

Ptolemy is often ridiculed for his geocentric model of the solar system, but he was mathematically brilliant. He computed an accurate model of the motions of the planets (*epicycles* - accurate from the earthly perspective, would just have to be re-calibrated every 100 years). Makes useful predictions given narrow domain of questioning. = Generalized system of approximation (Fourier series).

Linear regression is the geometric model of applied statistics. Linear regression: family of simple golems attempting to learn about the mean and variance of some measurement using an additive combinations of other measurements

### Linear regression as a Bayesian procedure

Which uses a **Gaussian (normal) distribution** to describe our golem's uncertainty about some measurement of interest.

**Binomial distribution** models the number of *discrete* successes (k) in a fixed number (n) of independent binary trials (Y/N, T/F). As the number of trials becomes large, the binomial distribution approaches the *continuous* Gaussian distribution, which is characterized by a bell-shaped curve defined by its mean and SD.

#### Normal by addition

Random walks on a soccer field.

```{r 4.1}
pos <- replicate(1000,sum(runif(16,-1,1)))
plot(density(pos))
```

INC number of steps INC stabilization on the Gaussian distribution. Any process that adds together random values from the same distribution converges to normal. Fluctuations from the mean cancel themselves out - most likely sum is sum in which every fluctuation is canceled by another (sum of zero)

-   Regardless of underlying distribution

#### Normal by multiplication 

Another way to get normal distribution.

```{r}
prod(1+runif(12,0,0.1))
```

-   Samples 12 random numbers between 1.0 and 1.1, each representing a proportional increase in growth. The product of all 12 is computed and returned.

```{r}
growth <- replicate(1e4, prod(1 + runif(12,0,0.1)))
dens(growth, norm.comp=TRUE)
```

The effect at each locus was multiplied by the effects at all the others, NOT ADDED. Normal distribution because the effect at each locus is SMALL (multiplying small numbers is approx same as addition).

::: callout-note
Small effects that multiply together are approximately additive, so they tend to stabilize on Gaussian distributions
:::

```{r 4.4}
big <- replicate(1e4, prod(1+runif(12,0,0.5)))
small <- replicate(1e4, prod(1 + runif(12,0,0.01)))
dens(big, norm.comp=TRUE)
dens(small, norm.comp=TRUE)
```

#### Normal by log-multiplication

```{r 4.5}
log.big <- replicate(1e4, log(prod(1+runif(12,0,0.5))))
dens(log.big, norm.comp=TRUE)
```

Adding logs is equivalent to multiplying the original numbers. Measurement scales are arbitrary, and it's natural to measure sound, earthquakes, information on a log scale.

### Using Gaussian distributions 

**Ontological reason:** World full of Gaussian distributions: measurement errors, variations in growth, velocities of molecules. \
Processes that add together fluctuations.

-   Member of **exponential family** of natural distributions

**Epistemological reason:** Maximum entropy → minimal assumptions about the data (just that the measure has finite variance)

```{r}
curve( exp( -x^2 ) , from=-3 , to=3 )
```

Gaussian is *continuous* - value $y$ can be any continuous value. In contrast to binomial, which requires integers.

-   Probability mass functions: discrete outcomes (like binomial) → denoted $Pr$

    -   Cannot be greater than 1

-   Probability density functions: continuous outcomes → denoted $P$ or $f$ .

    -   can be greater than 1

        -   Probability density is the rate of change in cumulative probability

        -   Area under the density function will never exceed 1 = probability mass

## A language for describing models 

::: {.callout-tip appearance="minimal"}
## Recipe for statistical models

1.  Recognize set of variables we want to understand. Data are observable variables. Parameters are unobservable, like rates and averages.
2.  For each variable, define it either in terms of other variables or a probability distribution
3.  Combination of variables and their probability distributions defines a *joint generative model* that can be used both to simulate hypothetical observations as well as analyze real ones.
:::

::: callout-note
## Summarizing the model

$y_{i}$ ~ $Normal(µ_{i}, σ)$

$µ{i} = 𝛽x{i}$

$𝛽$ \~ $Normal(0,10)$

$σ$ \~ $Exponential(1)$

$x_{i}$ \~ $Normal(0,1)$
:::

Revising the globe-tossing model:

$$
W \sim {\sf Binom}(N, p)
$$

$$
p \sim {\sf Uniform}(0,1)
$$

-   [Likelihood function used in Bayes' thm]{.underline}: The count $W$ is distributed binomially with sample size N and probability $p$

-   [Define priors]{.underline}: The prior for $p$ is assumed to be uniform between 0 and 1

**Stochastic**: indicated by \~ symbol. No single instance of the variable on left is known with certainty, but is instead characterized by a probabilistic mapping

```{r R code form}
w <- 6; n <- 9;
p_grid <- seq(0,1,length.out=100)
posterior <- dbinom(w,n,p_grid) * dunif(p_grid,0,1)
posterior <- posterior/sum(posterior)
```

## Gaussian Model of height

Loaded the data, not place them into convenient object

```{r 4.7}
d <- Howell1
```

-   **Data frame** simply named `d`: special kind of `list` in R, table with named columns, correponding with variables, and numbered rows, corresponding to individual cases

```{r 4.8}
# str() compactly display the structure of an R object 
str(d)

# precis() helpful for summarizing posterior distr further on 
precis(d)
```

Working with columns

-   `$` → extract the column named `height` from the data frame `d`

```{r 4.10}
#| output: false
d$height
```

```{r 4.11}
d2 <- d[d$age >= 18,]
```

-   Index notation: `[row, column]`

```{r}
dens(d2$height)
```

-   Height raw data looks approx Gaussian, BUT not a good idea to gawk at raw data and decide how to model them

Write general model and compute plausibility of each µ and σ. To define the heights as normally distributed with a mean µ and sd σ:

$$
h_{i} \sim {Normal}(µ,σ) 
$$

-   h → list of heights

-   subscript i –\> each individual element of list, takes on row numbers (1-32, no. heights in `d2$height` )

All the golem knows about each height measurement is defined by the name normal distribution with mean µ and standard dev σ. The model assumes that the values $h_{i}$ are **independent and identically distributed (IDD)**. In other words, each value of h has the same probability function, independent of the other h values and using the same parameters. But hardly true - heights within families are correlated.

#### Priors 

The parameters to be estimated are µ and σ → so need prior $Pr(µ, σ)$ or the joint prior probability for all parameters. Assume that they're independent of each other or $Pr(µ, σ) = Pr(µ)Pr(σ)$

$$
h_{i} \sim {Normal}(µ,σ) 
$$

$$
µ \sim{Normal}(178, 20)
$$

$$
σ \sim {Uniform}(0, 50)
$$

The prior for µ is a broad Gaussian prior, centered on 178cm, with 95% of probability between 178 ±40.

-   Why 178 cm? Author is 178 cm tall. And the range from 138 cm to 218 cm encompasses a huge range of plausible mean heights for human populations. So domain-specific information has gone into this prior. Everyone knows something about human height and can set a reasonable and vague prior of this kind.

Whatever the prior, it’s a very good idea to **plot your priors**, so you have a sense of the assumption they build into the model.

```{r}
curve(dnorm(x,178,20), from=100, to=250, xlab = "mu",ylab="density")
```

-   the golem is assuming that the average height (not each individual height) is almost certainly between 140 cm and 220 cm.

The σ prior is a truly flat prior, a uniform one, that functions just to constrain σ to have positive probability between zero and 50cm. View it with:

```{r}
curve(dunif(x,0,50), from=-10, to=60)
```

#### Prior predictive simulation

-   See what your priors imply about the distribution of heights

-   Helps diagnose bad choices

-   Simulate heights by sampling from the prior

```{r 4.14}
sample_mu <- rnorm(1e4, 178,20)
sample_sigma <- runif(1e4,0,50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h, xlab="height")
```

-   Expected distribution of heights, averaged over the prior

-   Prior prob distribution is not itself Gaussian - not an empirical expectation but a distribution of relative plausibilities of different heights, BEFORE seeing the data

Compare to a less informative prior for average height µ \~ Normal(178,100). Priors with large SDs are common in Bayesian models, but not really sensible.

```{r 4.15}
sample_mu <- rnorm(1e4, 178,100)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
abline(v=0, lty = "dashed") #plot line at x=0 to show where negative values occur
```

-   4% of expected heights are negative. Also expects some giants: 18% of expected heights over 272cm (tallest record)

#### Grid approximation of posterior distribution 

For the first time modeling Gaussian distribution, demonstrate calculations with brute force (usually impractical)

```{r guts of the Gaussian golem}
mu.list <- seq( from=150, to=160 , length.out=100 )
sigma.list <- seq( from=7 , to=9 , length.out=100 )
post <- expand.grid( mu=mu.list , sigma=sigma.list )
post$LL <- sapply( 1:nrow(post) , function(i) sum(
dnorm( d2$height , post$mu[i] , post$sigma[i] , log=TRUE ) ) )
post$prod <- post$LL + dnorm( post$mu , 178 , 20 , TRUE ) +
dunif( post$sigma , 0 , 50 , TRUE )
post$prob <- exp( post$prod - max(post$prod) )

contour_xyz(post$mu, post$sigma, post$prob)
image_xyz(post$mu, post$sigma, post$prob)
```

-   inspect the posterior distribution (in `post$prob` ) with functions like `contour_xyz()` or `image_xyz()` in the `rethinking` package

#### Sampling from the posterior 

To summarize and understand the posterior distribution, like in Ch3. New thing: since 2 parameters and need to sample combos of them, we first randomly samples row numbers in `post` in proportion to values in `post$prob`. Then pull out parameter values on those randomly sampled rows.

```{r 4.19}
sample.rows <- sample(1:nrow(post), size=1e4, replace=TRUE, prob=post$prob)
sample.mu <- post$mu[sample.rows]
sample.sigma <- post$sigma[sample.rows]

```

-   **Output**: 1e4 samples, with replacement, from the posterior for height data, plotted below:

```{r 4.20}
plot(sample.mu, sample.sigma, cex=0.5, pch=16, col=col.alpha(rangi2,0.1))
```

-   `col.alpha()` makes the colors transparent → demonstrates density (overlapping samples)

-   `cex` (character expansion, size of points) a numerical vector giving the amount by which plotting characters and symbols should be scaled relative to the default.

-   `pch` (plot character) a vector of plotting characters or symbols

Plot the densities of µ and σ → characterizing the shapes of the ***marginal*** **posterior densities of µ and** σ.

-   Marginal: averaging over the other parameters

```{r 4.21}
dens(sample.mu)
dens(sample.sigma)
```

```{r 4.22}
PI(sample.mu)
PI(sample.sigma)
```

Since these samples are just vectors of numbers, you can compute any statistic from them that you could from ordinary data.

**Sample size and the normality of the standard dev's posterior:** Posterior is not always so Gaussian in shape. For a Gaussian likelihood and a Gaussian prior on µ, the posterior distribution is always Gaussian as well, regardless of sample size. It is the standard deviation σ that causes problems. So if you care about σ—often people do not—you do need to be careful of abusing the quadratic approximation.

Posterior of σ has long right-hand tail. Variances (SDs) must be positive (more uncertainty about how big it is rather than how small it is.

Sample with small size of 20 heights:

```{r 4.23}
d3 <- sample(d2$height, size=20)

# repeat code from previous subsection
mu.list <- seq( from=150, to=170 , length.out=200 )
sigma.list <- seq( from=4 , to=20 , length.out=200 )
post2 <- expand.grid( mu=mu.list , sigma=sigma.list )
post2$LL <- sapply( 1:nrow(post2) , function(i)
sum( dnorm( d3 , mean=post2$mu[i] , sd=post2$sigma[i],log=TRUE ) ) )
post2$prod <- post2$LL + dnorm( post2$mu , 178 , 20 , TRUE ) +
dunif( post2$sigma , 0 , 50 , TRUE )
post2$prob <- exp( post2$prod - max(post2$prod) )
sample2.rows <- sample( 1:nrow(post2) , size=1e4 , replace=TRUE ,
prob=post2$prob )
sample2.mu <- post2$mu[ sample2.rows ]
sample2.sigma <- post2$sigma[ sample2.rows ]
plot( sample2.mu , sample2.sigma , cex=0.5 ,
col=col.alpha(rangi2,0.1) ,
xlab="mu" , ylab="sigma" , pch=16 )
```

Inspect the marginal posterior density for σ, averaging over µ, produced with:

```{r 4.25}
dens( sample2.sigma , norm.comp=TRUE )
```

-   the posterior for σ is not Gaussian, but rather has a long tail of uncertainty towards higher values

#### Finding posterior distribution with `quap`

**Quadratic approximation**: quickly make inferences about the shape of the posterior. The posterior's peak → *maximum a posteriori* estimate (MAP) = useful image of posterior's shape at this peak

`quap` in `rethinking` - define the posterior probability at each combination of parameter values.

-   Then climbs the posterior distribution and find the peak, its MAP.

-   Estimates the quadratic curvature at the MAP to produce an approximation of the posterior distribution.

-   Remember: This procedure is very similar to what many non-Bayesian procedures do, just without any priors.

```{r 4.26}
library(rethinking) # repeat code to load data and select out adults
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
```

| Model definition           | R code                     |
|----------------------------|----------------------------|
| $h_{i} \sim {Normal}(µ,σ)$ | height \~ dnorm(mu, sigma) |
| $µ \sim{Normal}(178, 20)$  | mu \~ dnorm(178,20)        |
| $σ \sim {Uniform}(0, 50)$  | sigma \~ dunif(0,50)       |

```{r 4.27}
flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(178, 20),
  sigma ~ dunif(0,50)
)
```

Fit model to the dataframe d2:

```{r 4.28}
# fit model stored in m4.1
m4.1 <- quap(flist, data=d2)

#look at posterior distribution 
precis(m4.1)
```

These numbers provide Gaussian approximations for each parameter’s *marginal* distribution. This means the plausibility of each value of µ, after averaging over the plausibilities of each value of σ, is given by a Gaussian distribution with mean 154.6 and standard deviation 0.4.

**Start values for quap** climbing a hill - needs to start at some combo of parameter values (usually random vals sampled from prior). Possible to specify starting val, like the mean and sd in this context.

```{r 4.30}
start <- list(
  mu=mean(d2$height),
  sigma=sd(d2$height)
)
m4.2 <- quap(flist, data=d2, start=start)
```

Difference between `alist` and `list`: both make collections of arbitrary R objects

-   `list` evaluates the code embedded inside - use when defining list of start values

-   `alist` does not execute code inside

Using a more informative prior (mean of 0.1 = narrow prior). µ won't change much, but the estimate for σ does

```{r 4.31}
m4.2 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ dnorm(178,0.1),
    sigma ~ dunif(0,50)
  ), data = d2
)
precis(m4.2)
```

Sampling from quap.

-   A quadratic approx to a posterior distribution with more than one parameter (e.g., µ and σ) is just a multi-dimensional Gaussian distribution

-   quap calculates deviations for all parameters and also **covariances** among pairs of parameters

    -   A list of means and a matrix of variances and covariances - sufficient to describe a multi-dimensional Gaussian distribution

**Variance-Covariance Matrix**: shows how each parameter relates to every other parameter in the posterior distribution.

```{r}
vcov(m4.1)
```

A variance-covariance matrix can be factored into two elements: (1) a vector of variances for the parameters and (2) a correlation matrix that tells us how changes in any parameter lead to correlated changes in the others.

```{r}
diag(vcov(m4.1)) # list of variances, if squared = the SDs in precis output 
cov2cor(vcov(m4.1)) # correlation matrix 
```

-   List of variances: if squared, we get the SDs in the `precis` output

-   Correlation matrix: correlations are close to 0, so learning µ tells us little about σ and vice versa

**Sample vectors of values from a multi-dimensional Gaussian distribution.** `extract.samples()` from the `rethinking` package.

-   `extract.samples` uses `mvrnorm()` which is like `rnorm` but simulates random vectors of multivariate Gaussian values.

```{r 4.34}
post <- extract.samples(m4.1, n=1e4)
head(post)
```

You end up with a data frame, post, with 10,000 (1e4) rows and two columns, one column for µ and one for σ. Each value is a sample from the posterior

```{r}
precis(post)

print("Vs m4.1 posterior")
precis(m4.1)

#Compare o the grid approx
plot(post)
```

## Linear Prediction 

**Predictor variable**: independent variable we posit predicts the outcome if it has statistical association with the outcome variable.

How Kalahari height (outcome var) covaries with weight (predictor var). Seems dull now because readers don't have a theory about growth and life history in mind.

Plot the height and weight against one another to get a sense of how they covary:

```{r 4.37}
plot(d2$height, d2$weight)
```

How to incorporate predictor vars into Gaussian model?

**The linear model strategy.** Make the parameter for the mean of a Gaussian distribution µ ito a linear function of the predictor variable and other new parameters we invent.

-   This instructs the golem to assume that the predictor variable has a constant and additive relationship to the mean of the outcome. It then computes the posterior distribution of this constant relationship

::: {.callout-note appearance="minimal"}
We ask the golem: “Consider all the lines that relate one variable to the other. Rank all of these lines by plausibility, given these data.”

The golem answers with a posterior distribution.
:::

Case study: one predictor variable (weight). How to get weight into Gaussian model of height? Define µ as a function of values in x (`d2$weight`), with $\overline{x}$ as the average of weights.

| $h_{i} \sim {Normal}(µ_{i},σ)$ | likelihood | `height ~ dnorm(mu,sigma)` |
|---------------------------|-----------------|----------------------------|
| $µ_{i} = \alpha + \beta(x_{i} - \overline x)$ | **linear model** | `mu ← a + b*weight` |
| $\alpha \sim{Normal}(178, 20)$ | $\alpha$ **prior** | `a ~ dnorm(178, 20)` |
| $\beta \sim{Normal}(0,10)$ | $\beta$ **prior** | `b ~ dlnorm(0,1)` |
| $σ \sim {Uniform}(0, 50)$ | $\sigma$ **prior** | `sigma ~ dunif(0,50)` |

#### Probability of the data

Probability of the observed height. The mean µ noe depends on unique values on each row i.

#### Linear Model

The mean µ is not longer a parameter to be estimated, instead $µ_{i}$ is determined by other parameters $\alpha$, $\beta$, and the observed variable $x$. We know $µ_{i}$ with certainty (not stochastic). $x_{i}$ is weight value on row $i$.

Asks two questions:

1.  What is the expected height when $x_{i} = \overline{x}$ ? $µ_{i} = \alpha$ when $x_{i} = \overline{x}$.
2.  What is the change in expected height when $x_{i}$ changes by 1 unit? $\beta$ is the slope - rate of change in expectation

Find the line that relates $x$ to $h$ - a line that passes through $\alpha$ when $x_{i} = \overline{x}$ and has slope $\beta$

#### Remaining priors 

These last priors define distributions for the unobserved variables.

$\beta$ prior: why Gaussian prior with mean 0?

Simulate a hundred pairs of alpha and beta:

```{r 4.38}
set.seed(2971) 
N <- 100      #100 lines
a <- rnorm(N,178,20)
b <- rnorm(N,0,10)
```

Now plot the lines:

```{r 4.39}
plot(NULL,xlim=range(d2$weight), ylim=c(-100,400), xlab="weight", ylab="height")

abline(h=0, lty=2) # no one is negative height
abline(h=272, lty=1, lwd=0.5) # tallest person 

mtext("b ~ dnorm(0,10")
xbar <- mean(d2$weight) 
for (i in 1:N) curve(a[i] + b[i] *(x-xbar),from = min(d2$weight),to=max(d2$weight), add=TRUE, col=col.alpha("black", 0.2))
```

-   Relationship between weight and height can be absurdly positive or negative

-   Know that height INC with avg weight. Restrict to positive values → define prior as $\beta \sim{Log-Normal}(0,1)$. Log of $\beta$ has a Normal(0,1) distribution

`dlnorm` and `rlnorm` densities for working with log-normal distributions

```{r 4.40}
b <- rlnorm(1e4, 0,1)
dens(b, xlim=c(0,5), adj=0.1)
```

-   If log of beta is normal, then beta itself is strictly positive

Re-doing the prior predictive simulation with Log-Normal

```{r 4.41}
set.seed(2971)
N <- 100 
a <- rnorm(N, 178,20)
b <- rlnorm(N,0,1)

# plotting
plot(NULL,xlim=range(d2$weight), ylim=c(-100,400), xlab="weight", ylab="height")

abline(h=0, lty=2) # no one is negative height
abline(h=272, lty=1, lwd=0.5) # tallest person 

mtext("log(b) ~ dnorm(0,1)")
xbar <- mean(d2$weight) 
for (i in 1:N) curve(a[i] + b[i] *(x-xbar),from = min(d2$weight),to=max(d2$weight), add=TRUE, col=col.alpha("black", 0.2))
```

-   Now all lines in the joint prior for alpha and beta are within human reason

Important to fuss over priors (even if they end up being irrelevant due to big data). Because some priors may end up being relevant for many other analyses and it helps us build better models.

::: callout-note
## Choosing priors

Priors encode states of info before seeing the data.

-   In cases with good prior info that discounts the plausibility of some parameter values (e.g., neg associations between height and weight) - encode directly in priors

-   Vary priors and repeat the analysis in order to study how different states of initial info influence inference

-   Often, many reasonable priors produce same inference

-   Conventional Bayesian priors are *conservative* relative to conventional non-Bayesian approaches

-   Choose priors conditional on **pre-data knowledge of the data (constraints, ranges, theoretical relationships**) never conditional on the observed sample
:::

#### Finding the posterior distribution

Just incorporate new model of the mean into the model specification inside `quap` and make sure to add a prior for the new parameter $\beta$

```{r 4.42}
#load data again, since it's a long way back 
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18,]

# define the average weight, x-bar 
xbar <- mean(d2$weight)

#fit model 
m4.3 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178,20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0,50)
  ),
  data=d2)
```

#### Interpreting the posterior distribution 

**Tables of marginal distributions**: inspect the marginal posterior distributions of the parameters

```{r 4.44}
precis(m4.3)
```

[Focus on new parameter $\beta$]{.underline}. Mean on 0.9 = *a person 1kg heavier is expected to by 0.90 cm taller*. 89% of the posterior probability lies between 0.84 and 0.97. $\beta$ vals close to 0 or greatly above 1 are highly incompatible with these data and this model.

-   Not suggesting that the relationship between height and weight are linear, just that: if you're committed to a line, then lines with slope `0.9` are plausible ones

Also need variance-covariance matrix to describe the quadratic posterior completely, along with `precis()`

```{r 4.45}
round(vcov(m4.3), 3)
```

Very little covariation among the parameters.

```{r}
pairs(m4.3) #shows both marginal posteriors and covariance 
```

**Plotting posterior inference against the data.**

Plot raw data, compute posterior mean values for a and b, draws implied line.

```{r 4.46}
plot(height ~ weight, data=d2, col=rangi2)
post <- extract.samples(m4.3)
a_map <- mean(post$a)
b_map <- mean(post$b)
curve(a_map + b_map * (x - xbar), add=TRUE)
```

Each point in this plot is a single individual. The black line (posterior mean) is defined by the mean slope β and mean intercept α. This is not a bad line. But lots of other highly plausible lines near it.

**Adding uncertainty around the mean**.

Each combo of $\alpha$ and $\beta$ has posterior probability. How to get uncertainty into the plot: sample a bunch of lines from the posterior distribution, display lines on plot to visualize the uncertainty in the regression relationship

Work with all the samples from the model

```{r 4.47}
post <- extract.samples(m4.3)
post[1:5,]
```

-   each row is a correlated random sample from the joint posterior of all three parameters, using the covariances provided by `vcov(m4.3)`

    -   The average of very many of these lines = **posterior mean line**

    -   Scatter around the average alters our confidence in the relationship between the predictor and outcome

Display bunch of the lines incrementally - see scatter

```{r 4.48}
N <- 100
dN <- d2[1:N,]
mN <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - mean(weight)),
    a ~ dnorm(178,20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0,50)
  ), data=dN
)

# Extract 20 samples from the posterior 
post <- extract.samples(mN, n=20)

# Display raw data and sample size 
plot(dN$weight, dN$height, 
     xlim=range(d2$weight), ylim=range(d2$height),
     col=rangi2, xlab="weight", ylab="height")
mtext(concat("N = ", N))

# Plot the lines, with transparency 
for (i in 1:20)
  curve(post$a[i] + post$b[i]*(x - mean(dN$weight)),
        col=col.alpha("black", 0.3), add=TRUE)
```

-   Notice that the cloud of regression lines becomes more compact as the sample size (N) increases. Model grows more confident about location of mean

**Plotting regression intervals and contours**: More informative to see uncertainty by plotting an interval or contour around the average regression line.

Make list of 10,000 vals of µ for an individual who weights 50kg ($x_{i}$) via samples from posterior

$$
µ_{i} = \alpha + \beta(x_{i}-\overline{x}) 
$$

```{r 4.50}
post <- extract.samples(m4.3)
mu_at_50 <- post$a + post$b * (50 - xbar)

# Plotting mu_at_50 = Vector of predicted means, one for each random sample from the posterior 
dens(mu_at_50, col=rangi2, lwd=2, xlab="mu|weight=50")
```

Since the components of µ have distributions, so too does µ. And since the distributions of α and β are Gaussian, so to is the distribution of µ (adding Gaussian distributions always produces a Gaussian distribution).

Since the posterior for µ is a distribution, you can find intervals for it, just like for any posterior distribution.

```{r 4.52}
PI(mu_at_50, prob=0.89)
```

-   the central 89% of the ways for the model to produce the data place the average height between about 159 cm and 160 cm (conditional on the model and data), assuming the weight is 50 kg.

That’s good so far, but we need to repeat the above calculation for every weight value on the horizontal axis, not just when it is 50 kg. Use the `link` function from `rethinking`. It takes the `quap` approx, samples from the posterior distribution, and computes µ for each case in the data and sample from the post distribution.

```{r 4.53}
mu <- link(m4.3)
str(mu)
```

-   Output: big matrix of µ values

    -   Each row = sample from post distri (default = 1000 samples)

    -   Each column = case in the data (352 columns in this matrix)

-   This is a distribution of µ for each individual in the original data

Now, let's get a distribution of µ for each unique weight value on the horizationtal axis:

```{r 4.54}
# Define sequence of wights to compute predictions for 
# these values will be on the horiz axis
weight.seq <- seq(25, 70, by=1)

# Use link() to compute mu 
# For each sample from posterior 
# and for each weight in weight.seq
mu <- link(m4.3, data=data.frame(weight=weight.seq))
str(mu)
```

-   Now only 46 columns in `mu` - fed it 46 diff values for `weight`

Visualize by plotting the distribution of µ values at each height

```{r 4.55}
# Use type="n" to hide raw data 
plot(height ~ weight, d2, type="n")

# Loop over samples and plot each mu value 
for (i in 1:100)
  points(weight.seq, mu[i,], pch=16, col=col.alpha(rangi2,0.1))
```

-   At each weight value in `weight.seq`, a pile of computed µ values are shown. Each of the piles is a Gaussian distribution

    -   The amount of uncertainty in µ depends on the value of `weight`

Summarize the distribution for each weight value, using `apply()`

```{r 4.56}
# Summarize the distribution of mu 

#Compute the mean of each column (dimension "2") of the matrix mu 
mu.mean <- apply(mu, 2, mean)   #contains the average mu at each weight value

#Compute the PI (89% interval) of each column of the matrix mu 
mu.PI <- apply(mu, 2, PI, prob=0.89)   #contains 89% lower and upper bounds for each weight value 
```

Plot these summaries on top of the data

```{r 4.57}
#plot raw data 
# fading out points to make line and interval more visible 
plot(height ~ weight, data=d2, col=col.alpha(rangi2,0.5))

# Plot the MAP line, aka the mu for each weight 
lines(weight.seq, mu.mean)

# Plot the shaded region for 89% PI 
shade(mu.PI, weight.seq)
```

Using this approach, can derive and plot the posterior prediction means and intervals for quite complicated models, for any data you choose.

::: callout-tip
## Recipe for generating predictions and intervals from the posterior of a fit model

1.  Use `link` to generate distributions of posterior values for µ. The default behavior of `link` is to use the original data, so you have to pass it a list of new horiz axis values you want to plot posterior predictions across
2.  Use summary functions like `mean` or `PI` to find averages and lower and upper bounds of µ for each value of the predictor variable
3.  Finally, use plotting functions like `lines` or `shade` to draw the lines and intervals. Or you might plot the distributions of the predictions, or do further numerical calculations with them

Recipe words for every model we fit in the book. As long as you know how parameters relate to the data, you can use samples from the posterior to describe any aspect of the model's behavior.
:::

Prediction intervals - for actual heights, not just mean heights $h_{i} \sim {Normal}(µ_{i},σ)$

-   Spread around µ govered by $\sigma$ → need to incorporate into predictions

-   `sim()` → Simulating heights: for each unique weight value, sample from a Gaussian distribution with the correct mean µ for the weight and correct value of SD $\sigma$ from the same posterior distribution. Simulation embodies the uncertainty in the posterior AND the Gaussian distribution of heights

```{r 4.59}
sim.height <- sim(m4.3, data=list(weight=weight.seq))
str(sim.height)
```

-   This matrix contains simulated heights, NOT distributions of plausible avg height, µ

Summarize the simulated heights using `apply`

```{r 4.60}
# Assin the 89% posterior prediction interval of observable (according to the model) heights, across the values of weight in weight.seq 
height.PI <- apply(sim.height, 2, PI, prob=0.89)
```

```{r 4.61}
# Plot raw data 
plot(height ~ weight, d2, col=col.alpha(rangi2, 0.5))

# Draw MAP line 
lines(weight.seq, mu.mean)

# Draw HPDI region for line 
mu.HPDI <- apply(mu, 2, HPDI, 0.89) # not sure about this..
shade(mu.HPDI, weight.seq)

# Draw PI region for simulated heights
shade(height.PI, weight.seq)
```

-   89% prediction interval for height as function of weight.

    -   **Solid line** = average line for the mean height at each weight

    -   **Narrow shaded interval** = distribution of µ

    -   **Wider shaded region** = region within which the model expects to find 89% of the actual heights in the population, at each weight

-   Uncertainty in both the parameter values AND in the sampling process
