---
title: "SR_Lec03_notes"
format: html
editor: visual
---

# LEC 03: Geocentric Models Notes

```{r setup, include=FALSE}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[d$age >= 18,]
```

Why normal?

1.  generative: summed fluctuations tend toward normal distribution
2.  inferential: for estimating mean and variance, normal distribution is least informative distribution (max entropy → fewest assumptions)

## Step 1: Question/goal/estimand

-   describe association between adult weight and height

## Step 2: Scientific model

How does height causally influence weight?

-   DAG:

$$
H → W ← U
$$

$$
W = f(H, U)
$$

-   "Weight is some function of height and unobserved stuff"

### Generative models

Options

1.  **Dynamic**: incremental growth of organism; both mass and height (length) derive from growth pattern; Gaussian variation result of summed fluctuations
2.  **Static**: (what we're doing) imagine association between variables w and h at particular age ranges but no mechanism, still at Gaussian variation, imagined as a result of growth history (but not modeled)

For adults, weight is a proportion of height plus the influence of unobserved causes:

$$
 W = \beta H + U
$$

-   beta is some proportion

### **Generative code**

```{r}
# function to simulate weights of individuals from height 
sim_weight <- function(H,b, sd) { 
  "simulate U using rnorm for each individual in H, using mean of 0 and sd = standard 
  deviation of U, unobserved noise"
  U <- rnorm(length(H), 0, sd) 
  W <- b*H + U
  return(W)
}
```

-   Goal - estimate $\beta$ but right now, simulating what data look like for any given value of beta

"This is the generative model: can it produce data that looks anything like the original data?" →

```{r}
H <- runif(200, min=130, max=170) # why uniform? just cuz he wanted the full x-axis represented
#H <- rnorm(200, 170,20)
#plot(H)

W <- sim_weight(H, b=0.5, sd=5)
plot(W~H, col=2, lwd=3)
```

-   Yes, it can (doesn't mean it's the right generative model)
-   Simulated individuals → biologically realistic relationships between weight and height

## Step 3: Statistical model(s)

::: callout-note
## Conventional statistical model notation

\(1\) List the variables

\(2\) Define each variable as a deterministic or distributional function of other variables
:::

|  |  |
|----|----|
| $W_{i} = \beta H_{i} + U_{i}$ | Equation for expected weight |
| $U_{i} \sim {\sf Normal}(0, \sigma)$ | Gaussian error with standard deviation sigma |
| $H_{i} \sim {\sf Uniform}(130,170)$ | Height uniformly distributed from 130-170cm |

### Estimator

We want to estimate how the avg weight changes with height

$$
E(W_{i}|H_{i}) = \alpha + \beta H_{i}
$$

-   $E(W_{i}|H_{i})$ = average weight conditional on height

How this relates to posterior distributions:

$$
Pr(\alpha,\beta,\sigma|H_{i},W_{i})=\frac{Pr(W_{i}|H_{i},\alpha,\beta,\sigma)Pr(\alpha,\beta,\sigma)}{Z}
$$

-   $Pr(\alpha,\beta,\sigma|H_{i},W_{i})$ → posterior probability of specific line

-   $Pr(W_{i}|H_{i},\alpha,\beta,\sigma)$ → Garden of forking data

-   $Pr(\alpha,\beta,\sigma)$ → prior

-   $Z$ → normalizing

### **Quadratic approximation**

-   Approximate the posterior distribution as a multivariate Gaussian distribution

<!-- -->

-   Posterior distributions are very often multivariate Gaussian distributions if the sample is reasonably large

    ::: callout-caution
    Quadratic approximation does almost what **Maximum Likelihood Estimation (non-Bayesian)** does in almost all cases.
    :::

-   `quap` tool so that students enter a list of deterministic and distributional assumptions that define a statistical model - runs it and finds the quadratic approximation of the posterior distribution

In statistical model, UNLIKE generative model, you need priors (when there's no observations, what does the model believe?)

-   Conventionally: you could use flat priors

-   Recommended: concentrated priors, these do NOT build in the answer but in fact, build AWAY from the learning from the data intentionally .

    -   Constrain implied observations of the model to scientifically realistic ranges

### Prior predictive distribution → same as posterior predictive distributions (same animal, same golem)

-   Priors should express scientific knowledge, but **softly**

    -   don't want predictions that mimic our sample because we haven't looked at our sample yet, we want boundaries that are realistic and reasonable

    -   Often generative process in nature is different than we imagine → want stat model to see and learn about those violations

    -   Scientific knowledge in this case: human biology...

+-------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------+
| Scientific knowledge (prior)                                                                                | Constraint explanation                                                                                                                                                  | Prior distributions               |
+=============================================================================================================+=========================================================================================================================================================================+===================================+
| When H = 0, W = 0 (someone with 0 height, should also weight 0 kg)                                          | Strong constraint on alpha (intercept): should be close to 0, so give wide (soft) variance: $Normal(0,10)$                                                              | $\alpha \sim {\sf Normal}(0,10)$  |
|                                                                                                             |                                                                                                                                                                         |                                   |
|                                                                                                             | alpha is the expected value of weight when height=0                                                                                                                     |                                   |
+-------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------+
| Weight INC (on average) with height (POSITIVE); Weight (kg) is less than height (cm) (BETWEEN ZERO AND ONE) | Constraint on slope: beta should be positive, between 0 and 1: $Uniform(0,1)$                                                                                           | $\beta \sim {\sf Uniform}(0,1)$   |
|                                                                                                             |                                                                                                                                                                         |                                   |
|                                                                                                             | Not the best choice, but for pedagogical reasons. Typically use Gaussian priors on slopes.                                                                              |                                   |
+-------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------+
| Standard deviations are positive                                                                            | Sigma must be positive → greater than 0, SD of 10 is variance of 100 kg and there's not a variance of 100 kg around avg weight so must be in this range: $Uniform(0,1)$ | $\sigma \sim {\sf Uniform}(0,10)$ |
+-------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------+

::: callout-important
## Normal VS Uniform Distribution

**Normal distributions**:

-   Bell-shaped: data clusters around a central mean

-   Model natural phenomena

**Uniform distributions**:

-   Rectangular - all outcomes are equally likely across a range

-   Model consistent, equal probability scenarios
:::

### Perform prior predictive **simulation**

-   Draw the observable implications of these priors in the observable outcome space. I.e., make some lines

```{r}
n <- 1e3 # num simulations 
a <- rnorm(n, 0,10) # draw n random alpha from prior distribution
b <- runif(n, 0,1) # draw n random beta from prior distribution

# Plot implied lines (50), set NULL plot has no points, 
plot(NULL, xlim=c(130,170), ylim=c(30,70), xlab="height (cm)", ylab="weight (kg)")
for (j in 1:50) abline(a=a[j], b=b[j], lwd=2, col=2)
```

This is what the golem imagines before seeing any data... a bit wild.

-   Constraint on beta from 0 to 1 is reasonable - so the slopes are not bad.

-   However, many lines are too low... impossibly so. Someone who is 170cm is not gonna be, on avg 30kg. High lines are impossibly high.

To see this prior predictive distribution (and the implications of your assumptions), as an expert in your field, in this visual way - you realize your assumptions are bad

-   Leave these priors for now. Won't do damage in this simple model. Bayesian is gonna learn the proper relationships among the variables quickly right away.

    ::: callout-note
    ## Sermon on priors

    -   FALSE ASSUMPTION: Bayesian stats is frequentism + priors

    <!-- -->

    -   **Priors are just posterior distributions. It's the posterior distribution, and how it's computed, that's diff from frequentist stats**

    -   There are no correct priors, only scientifically justifiable priors (same with posteriors)

        -   Scientific assumptions justify the generative model, which justify the estimator

        -   Justify with info outside the data - like the rest of model

        -   Priors are not so important in simple models

        -   Very important/useful in complex models

            -   Need to practice not: simulate, understand

    -   Easy to do better than "flat" priors. If you don't **regularize**, then you are exaggerating effect sizes.

    -   Anti-Bayesian boomers worry priors are used to cheat, but priors nearly always used to *shrink effects to zero!* Bayes is **conservative**.
    :::

## Step 4: Validate model

### Simulation-based validation

-   Bare minimum: test statistical model with simulated obs from scientific model

-   Golem might be broken

-   But regardless of what package you're using (like `quap`), even working golems might not deliver what you hoped

    -   Generative process - write it as code to generate synthetic data to show that your estimator, whatever structure it has, can work even in principle

-   Stronger test: **Simulation-based calibration**

    -   More stringent tests, how calibrated Bayesian estimators are over full shape

When validating your model with simulations

-   Vary slope and make sure posterior mean tracks it

-   Use large sample to see that it converges to data-generating value

-   Same for other unknowns (parameters)

```{r simulate}
# Simulate a sample of 10 people 
set.seed(93)
H <- runif(10, 130, 170)
W <- sim_weight(H, b=0.5, sd=5) #data-generating slope will be 0.5, sd=5 or var of 25kg

# run the model to compare
library(rethinking)
m3.1 <- quap( 
  alist(        #quap takes a formula list that mirrors the stat model
    W ~ dnorm(mu, sigma), 
    mu <- a + b*H,
    a ~ dnorm(0,10),
    b ~ dunif(0,1),
    sigma ~ dunif(0,10)
  ), data=list(W=W, H=H))

# summary 
precis(m3.1)
```

-   As you see in the precis: the model converges to the data-generating slope $\beta$ we gave, as well as the standard deviation $\sigma$

    -   But having trouble with the intercept $\alpha$

## Step 5: Analyze data

After tiresome rounds of model validation (takes very long with more complex models), you can take real data sample and the model will run.

Take Howell's data and use it in your model

```{r analysis}
dat <- list(W=d2$weight, H=d2$height)
m3.2 <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + b*H,
    a ~ dnorm(0,10),
    b ~ dunif(0,1),
    sigma ~ dunif(0,10)
  ), data=dat
)
precis(m3.2)
```

-   beta is around 0.5 yay

-   sigma is around 5

-   but intercept alpha is definitely not zero → suggests that there's something globally nonlinear about the relationship between height and weight (cuz you can't have negative weights)

    -   But works for ADULT w and h

    -   This intercept is just a weird, non-biological term that's needed to make the line fit

**Pairs plot** plots the pairs of unknowns from the posterior distrib against one another

```{r pairs}
pairs(m3.2)
```

-   [Diagonal]{.underline}: density of each unknown (a,b, sigma) → Gaussian bc that's what quap produces

-   [Upper right 3 cells]{.underline}: Posterior distributions as plotted from above for each pair

    -   Intercept and slope are very correlated - strong covariation along tight line (how linear regressions often work)

    -   Can't interpret intercept and slope separately

#### First Law of statistical interpretation:

-   the parameters are not independent of one another and cannot be independently interpreted

-   Instead, push out posterior predictions and describe/interpret those

    -   `extract.samples` creates synthetic data table of samples from the posterior distributions that adhere to the covariance structure

### Posterior predictive distribution

**The posterior is full of lines**

```{r}
# extract samples from the model of weight as a function of height for the actual Howell sample 
post <- extract.samples(m3.2)

# Plot the Howell sample, each red dot is person in Howell's dataset 
plot(d2$height, d2$weight, col=2, lwd=3, xlab="height (cm)", ylab="weight (kg)")

# Draw 20 lines randomly sampled from the posterior distributions
for (j in 1:20)
  abline(a=post$a[j], b=post$b[j], lwd=1)
```

Bayesian updating perspective

-   ... more red dots - more narrow range of lines

**The posterior is full of people**

-   Lines of posterior - slopes very little bit but in a fairly narrow range

    -   Estimates ONLY the AVERAGE relationship between weight and height

-   Most of red dots are not on the line → Lots of variation from the U (unobserved stuff)

Sometimes you want the U on the graph, which comes from sigma

```{r}
# Plot the Howell sample, each red dot is person in Howell's dataset 
plot(d2$height, d2$weight, col=2, lwd=3, xlab="height (cm)", ylab="weight (kg)")

# Draw 20 lines randomly sampled from the posterior distributions
for (j in 1:20)
  abline(a=post$a[j], b=post$b[j], lwd=1)

# put some new data ranging between 130-190 cm
height_seq <- seq(130, 190, len=20) 

# use sim, give it the fit quap model and the new data 
W_postpred <- sim(m3.2,
                  data=list(H=height_seq))

#take percentile interval for each height value 
W_PI <- apply(W_postpred, 2, PI)
lines(height_seq, W_PI[1,], lty=2, lwd=2)
lines(height_seq, W_PI[2,], lty=2, lwd=2)
```
