---
title: "distributions explainer"
format: html
editor: visual
execute:
  eval: false
---

## Distributions: Improving your priors

::: callout-important
## My questions about Bayesian workflow

To make predictions with a Bayesian model, is it okay to look at the mean of the raw data to set my priors? Or is it better not to look at the raw data's summary?

> Looking at the mean of the predictor variables is generally fine, especially while you're learning. The important distinction is:
>
> ✅ Using data summaries to set weakly informative priors (reasonable boundaries)
>
> ❌ Using data summaries to set very tight priors that essentially "peek" at the answer
:::

## Log-Normal Distribution

```{r}
# | echo: false
# These two are identical:
rlnorm(N, 0, 1)
exp(rnorm(N, 0, 1))
```

### Practical Intuition

Since `rnorm(N, 0, 1)` gives values mostly between **-2 and +2**, after `exp()` that becomes roughly **0.14 to 7.4** — a huge range for a slope

> $e^2 \approx 7.4$

With `rnorm(N, 0, 0.5)`, values are mostly between **-1 and +1**, so after `exp()` you get roughly **0.37 to 2.7** — much more controlled.

> $e^2 \approx 2.7$

### Sample and Plot

```{r}
# Try different SD values and see what slopes you get
b1 <- rlnorm(1000, 0, 1)    # original
b2 <- rlnorm(1000, 0, 0.5)  # tighter
b3 <- rlnorm(1000, 0, 0.25) # very tight

par(mfrow=c(1,3))
hist(b1, main="SD=1", xlim=c(0,10))
hist(b2, main="SD=0.5", xlim=c(0,10))
hist(b3, main="SD=0.25", xlim=c(0,10))
```

Then ask yourself: **does a slope of 7 make sense?** (Ch4.7 Hard Qs example: That would mean every 1 kg of weight adds 7 cm of height — probably too steep for children)
