---
title: "Rethinking Cheatsheet"
editor: visual
execute:
  eval: false
---

`extract.prior`

> **Action**: simulates draws using the defined formulas and prior specifications
>
> **Target**: *prior* distribution
>
> **Purpose**: to perform prior predictive simulation, checking what the model assumes before it looks at real data to ensure priors are not absurd

`extract.samples`

> **Action**: collects samples from a fitted model object (like `quap`)
>
> **Target**: posterior distribution
>
> **Purpose**: analyze parameter estimates, compute contrasts, run posterior predictive checks after learning from data

`link` function (`rethinking` package)

> Takes the `quap` approx, samples from the posterior distribution, and computes µ for each case in the data and sample from the post distribution.
>
> Default behavior: uses original data, so you have to pass it a list of new x-axis values to plot the posterior predictions across (`predictor_seq`)

```{r}
predictor_seq <- seq(from=X, to=XX, length.out=30) 
link(quap_model, post=prior, data=data.frame(predictor_var = predictor_seq))
```

::: callout-tip
## Workflow for generating predictions and intervals from the posterior of a fit model

1.   `link` to generate distributions of posterior values for µ.
    - Pass it a list of new horiz axis values you want to plot posterior predictions across
2.  Summary functions (`mean` or `PI`) to find averages and lower and upper bounds of µ for each value of the predictor variable
3.  Plotting functions (`lines` or `shade)` to draw the lines and intervals.
    - Or you might plot the distributions of the predictions, or do further numerical calculations with them
:::

Categorical Linear Regression

> **Purpose**: to build models with different [intercepts]{.underline} $\alpha$ for categorical predictor variables.
>
> Can use indicator (0/1) variables or index variables (ID=\[1,2,...,k\]). McElreath recommends index variables.

------------------------------------------------------------------------

## Quarto Cheatsheet 

#### Code Fold

``` R
#| code-fold: true
#| code-summary: "Click to see the R code"
```

```{r}
#| code-fold: true
#| code-summary: "Click to see the R code"

"some code here which is hidden, but the output is shown"

```

#### Hide source code while still executing it

``` R
#| echo: false
```

#### Make Output Plot's Background Transparent

``` R
#| dev: png
#| dev.args: !expr list(bg = "transparent")
```
